---
title: "Microresp pipeline"
author: "David Rodrigo Cajas"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
  html_document:
    toc: true
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 2*1920/300, fig.height = 2*1080/300, dpi = 300)
```

## 0) Data import

### 0.1) Working directory and packages

First we need to install the Rstudioapi package

```{r}
# (Install and) load Rstudio api package
if ("rstudioapi" %in% installed.packages()) {
  library(rstudioapi)
} else {
  install.packages("rstudioapi")
  library(rstudioapi)
} 
```

Now we will set the working directory to where the R markdown folder is

```{r}
wd <- dirname(rstudioapi::getSourceEditorContext()$path)

setwd(wd)
```

Import the list of packages in "required_packages.rds" and install them

```{r}
# Load list of required packages
required_packages <- readRDS("required_packages.rds")
# Install script's required packages
need_install <- required_packages[!(required_packages) %in% installed.packages()]
if (length(need_install) > 0) {
  install.packages(need_install)
}
```

[Later on] If you modified this code, don't forget to update the list of script's required packages

```{r}
# required_packages <- names(sessionInfo()$otherPkgs)
# saveRDS(required_packages, "required_packages.rds")
```

### 0.2) Import measurements data

Import the first page inside the "pathogen_growth_inhibition.xlsx" excel file.

```{r}
library(readxl)
library(dplyr)
library(stringr)
library(gtools)
library(tidyverse)

# set the file containing the data
sourcefile <- "pathogen_growth_inhibition.xlsx"

results <- read_excel(sourcefile, sheet = 1) # import columns
results <- drop_na(results,"Sample") # Remove rows where the Sample value is NA

# Set the right type for every column
results$Week <- as.factor(results$Week)
results$Sample <- as.factor(results$Sample)
results$Agrobacterium_colonies_5dpi_B_relative_growth <- as.factor(results$Agrobacterium_colonies_5dpi_B_relative_growth)

# See the dataframe
(results)
```

### 0.3) Import samples metadata

#### 0.3.1) Import metadata from samples

```{r}
library(readxl)
library(stringr)
library(tidyr)

# Import sample labels from Metadata page in "pathogen_growth_inhibition.xlsx" spreadsheet
meta <- read_excel(sourcefile, sheet = "Metadata") # import columns
meta$Sample <- as.factor(meta$Sample) # Make sure the Sample column is of type character
meta$replicate <- as.factor(meta$replicate)

meta

```

#### 0.3.2) Import experiment metadata

Note that this script imports the metadata from "experiment_metadata.xlsx" assuming that it's separated by sheets called "treatment_data", "soil_data" and "plant_data". The order is not relevant.

```{r}
library(readxl)

# Define source of experiment metadata

sourcemeta <- "experiment_metadata.xlsx"

for (i in excel_sheets(sourcemeta)) {
  # Import data
  
  df <- read_excel(sourcemeta, sheet = i) # import all experiment metadata in separated dataframes per sheet
  
  # Wrap up
  
  assign(i, df) # name output dataframe
  rm(df,i) # remove auxiliary "df" and "i" objects
}

# Add order to treatments
treatment_data$label <- factor(as.factor(treatment_data$label), levels = c("Control", "Disease suppression", "AMF", "Nitrogen fixation", "Phosphate solubilisation"))
treatment_data$applied_product <- factor(as.factor(treatment_data$applied_product), levels = c("No product", "Compete Plus", "MycorGran 2.0", "Vixeran", "NuelloPhos"))

# Small processing of soil data

numeric_cols <- names(soil_data)[sapply(soil_data, function(x) any(grepl("[0-9]", x)))] # Auxiliary object listing the columns that contain numbers
soil_data[numeric_cols] <- lapply(soil_data[numeric_cols], function(x) {
  x <- ifelse(grepl("^<", x), 0, x) # Identify values beginning with "<", which are below detection limitm and replace them with 0
  x <- as.numeric(x)  # Convert to numeric AFTER cleaning
  x
}) # This function replaces values that start with "<" with 0
rm(numeric_cols) # Remove auxiliary object
```

#### 0.3.3) Extract relevant experiment metadata and merge it into our metadata spreadsheet

```{r}
library(dplyr)

# Define relevant metadata to extract. This can be modified for further customisation
pick_metadata <- c(colnames(soil_data[,c(3,6,10:14)]), # chosen metadata columns in soil_data
                   colnames(treatment_data[,c(3,5:14)]), # chosen metadata columns in treatment_data
                   colnames(plant_data[,c()])) # chosen metadata columns in plant_data

# Create a function to replace values in a common column. It will be used to replace labels
replace_values <- function(df1, df2, col_name, new_col_name) {
  df1 %>%
    left_join(df2, by = col_name) %>%  # Join with lookup table
    mutate(!!col_name := !!sym(new_col_name)) %>% # Replace original values with labels
    select(-!!sym(new_col_name)) # Delete the new_col_name column can now be deleted
} # replaces the values of the col_name column in the df1 for the values of the new_col_name in the df2, assuming col_name exist in both dataframes

# Add selected metadata and change labels to metadata dataframe

  meta <- meta %>%
    
    # Add defined metadata
    
    left_join(select(plant_data, "plant",any_of(pick_metadata)), by = "plant") %>%
    left_join(select(soil_data, "soil",any_of(pick_metadata)), by = "soil") %>%
    left_join(select(treatment_data, "treatment",any_of(pick_metadata)), by = "treatment") %>%
    
    # Replace compressed labels for full size variables
    
    replace_values(plant_data[,1:2], "plant", "label") %>%
    replace_values(soil_data[,1:2], "soil", "label") %>%
    replace_values(treatment_data[,1:2], "treatment", "label")

meta
```

#### 0.3.4) Paste metadata dataframe from 0.3.3) into Results dataframe imported in 0.2)

```{r}
library(dplyr)

results <- left_join(results,meta,by = "Sample")
results
```

## 1) Data processing

Some calculations will be done to further process the data:

-   Fungal growth rate: $Growth~rate~(mm/day) = \frac{colony~radius~(mm)}{time~(days)}$

-   Inhibition rate (based on [Sarven MS et al. 2020](https://www.mdpi.com/2076-0817/9/3/213). doi 10.3390/pathogens9030213): $Inhibition~rate~(\%) = \frac{Growth~rate~(mm/day)_{Sterile} - Growth~rate~(mm/day)_{Sample}}{Growth~rate~(mm/day)_{Sterile}} \times 100$

```{r}
# Calculate growth rates
results$Fusarium_growth_rate_A_sterile_mmday <- (results$Fusarium_growth_6dpi_A_sterile_mm/6)

results$Fusarium_growth_rate_B_mmday <- (results$Fusarium_growth_14dpi_B_mm/14) # taking the 14 dpi data as a measurement point

# Calculate inhibition growth rates

results$Fusarium_inhibition_rate <- (100*(results$Fusarium_growth_rate_A_sterile_mmday - results$Fusarium_growth_rate_B_mmday) /  results$Fusarium_growth_rate_A_sterile_mmday)

```

Let's check a simplified view of the dataset

```{r}
interest_columns <- c("Sample", "soil", "treatment", "Fusarium_growth_14dpi_B_mm", "Fusarium_inhibition_rate")
select(results, interest_columns)
```

## 2) Data visualization

Just for visual consistency. A standard color palette

```{r}
# 5 color palette 
palette_5col <- c("#EB8900", "#47D7AC", "#FAD847", "#2761C4", "#962C5D", "#CC4389")
# palette_5col <- c("#EB8900", "#2761C4", "#47D7AC", "#CC4389", "#FAD847")

# Treatment palettes based on the 5 color palette
palette_treatments <- palette_5col
names(palette_treatments) <- levels(treatment_data$label)

palette_products <- palette_5col
names(palette_products) <- levels(treatment_data$applied_product)
```

Plot both the raw measurements along with the inhibition rate indicator

```{r}
library(ggplot2)
library(ggpubr)
library(ggprism)
library(plotly)

plot_fusarium_growth_sterile <- ggplot(drop_na(drop_na(results,"treatment"),"Fusarium_growth_6dpi_A_sterile_mm")
       , aes(x = soil, y = Fusarium_growth_6dpi_A_sterile_mm, color = treatment)) +
  geom_point() +
  geom_boxplot() +
  scale_color_manual(values = palette_treatments) +
  theme_prism() + 
  labs(x = "Soil", y = "Growth (mm)") +
  ggtitle("F. oxysporum colony radius in autoclaved plates 6 days post-inoculation")

plot_fusarium_growth_sample <- ggplot(drop_na(results,"Fusarium_growth_14dpi_B_mm")
       , aes(x = soil, y = Fusarium_growth_14dpi_B_mm, color = treatment)) +
  geom_point() +
  geom_boxplot() +
  scale_color_manual(values = palette_treatments) +
  theme_prism() + 
  labs(x = "Soil", y = "Growth (mm)") +
  ggtitle("F. oxysporum colony radius in non-autoclaved plates 14 days post-inoculation")

plot_fusarium_inhibition <- ggplot(drop_na(results,"Fusarium_inhibition_rate")
       , aes(x = soil, y = Fusarium_inhibition_rate, color = treatment)) +
  geom_point() +
  geom_boxplot() +
  scale_color_manual(values = palette_treatments) +
  theme_prism() + 
  annotate("text",
           x = 1.5, # Midpoint of the x-axis, adjust if you have more soil types
           y = 90, # Place it slightly above the max data point. Adjust this value.
           label = expression(paste("Inhibition rate (%) = ", 100 * frac(Growth~rate~(mm/day)[Sterile] - Growth~rate~(mm/day)[Sample], Growth~rate~(mm/day)[Sterile]))),
           hjust = 0, # Center horizontally
           vjust = 0,   # Align the top of the text with the y-coordinate
           size = 3) +
  coord_cartesian(clip = "off") + # Allows text to go outside the standard plot area
  labs(x = "Soil", y = "Rhizosphere inhibitory effect (%)") + ggtitle("F. oxysporum growth inhibition by soil samples")

ggarrange(plot_fusarium_growth_sterile,plot_fusarium_growth_sample, nrow = 1, common.legend = T)
plot_fusarium_growth_sterile
plot_fusarium_growth_sample
plot_fusarium_inhibition
```

## 3) Statistics

### 3.1) Fit a model to the data

Finding de model that best describes the data:

```{r}
library(lme4)
library(nlme)
library(lmerTest)
library(pbkrtest)

# Using lmer() function

model_random_a <- lmer(Fusarium_inhibition_rate ~ soil + treatment + soil:treatment + (1|replicate), data = drop_na(results,"Fusarium_growth_rate_B_mmday")) # Mixed model with soil, treatment and their interaction as fixed effect, plus a random replicate effect


model_nest_a <- lmer(Fusarium_inhibition_rate ~ soil + treatment + soil:treatment + (1|treatment:replicate), data = drop_na(results,"Fusarium_growth_rate_B_mmday")) # Same as before, but the replicate effect is nested inside the treatment.

# Using lme() function

model_random_b <- lme(Fusarium_inhibition_rate ~ soil + treatment + soil:treatment, random = ~ 1|replicate, data = drop_na(results,"Fusarium_growth_rate_B_mmday")) # Mixed model with soil, treatment and their interaction as fixed effect, plus a random replicate effect

model_nest_b <- lme(Fusarium_inhibition_rate ~ soil + treatment + soil:treatment, random = ~ 1|soil/replicate, data = drop_na(results,"Fusarium_growth_rate_B_mmday")) # Same as before, but the replicate effect is nested inside the treatment.

# Using lm() function

model_fixed_a <- lm(Fusarium_inhibition_rate ~ soil/replicate + soil + treatment + soil:treatment, data = drop_na(results,"Fusarium_growth_rate_B_mmday")) # Linear model with soil, treatment, their interaction and the replicate effect all defined as fixed factors.

model_fixed_b <- lm(Fusarium_inhibition_rate ~ soil + treatment + soil:treatment, data = drop_na(results,"Fusarium_growth_rate_B_mmday")) # Linear model with soil, treatment and their interactions as fixed factors, disregarding replicate effect entirely.


### The mixed models with no nesting of the replicate effect were more successful on fitting the data, therefore explain better the dataset and will be used in further analyses.

summary(model_random_a)
# summary(model_random_b)
# summary(model_fixed_a)
# summary(model_fixed_b)

anova(model_random_a)
# anova.lme(model_random_b)
# anova(model_fixed_a)
# anova(model_fixed_b)


```

The mixed linear model suggests there is are inconsistent soil effects and treatment effects. The ANOVA does not find significant soil or treatment effect but it does show a significant soil:treatment interaction. The comparison will be carried then in an interaction cohort basis, ie: treatment difference within soils.

### 3.2) Run multiple comparisons test

```{r}
library(emmeans)
# library(multcomp)
library(dplyr)

# Get estimated marginal means for the 'group' factor
paircomp <- emmeans(model_random_a, pairwise ~ treatment|soil, adjust = "tukey")
# (paircomp <- emmeans(model_random_a, pairwise ~ treatment|soil_texture_main, adjust = "tukey"))

significant <- as.data.frame(paircomp$contrasts)[as.data.frame(paircomp$contrasts)$p.value < 0.1, ]
significant <- separate(significant, contrast, into = c("treatment1", "treatment2"), sep = " - ")

significant

```

According to the multiple comparisons analysis, the only soil where the treatment produces a significant change is in the Clay soil. The additional sample measurement for the Sandy high Phosphate soil might change the result for this comparison.

### 3.3) Add statistics to the original plot

```{r}

significant <- significant %>%
  group_by(soil) %>%
  mutate(y_position = max(as.data.frame(paircomp)$upper.CL) * 1.05 + 
                        (row_number() - 1) * max(as.data.frame(paircomp)$upper.CL) * 0.1) %>%
  ungroup()

ggplot(drop_na(results,"Fusarium_inhibition_rate")
       , aes(x = treatment, y = Fusarium_inhibition_rate, color = treatment)) +
  facet_wrap(~soil, nrow = 1) +
  geom_boxplot() +
  scale_color_manual(values = palette_treatments) +
  theme_prism() +
  coord_cartesian(clip = "off") + # Allows text to go outside the standard plot area
  theme(axis.text.x = element_blank()) + 
  labs(x = "", y = "Rhizosphere inhibitory effect (%)") + ggtitle("F. oxysporum growth inhibition by soil samples") + 
  geom_signif(
  data = significant,
  aes(xmin = treatment1, xmax = treatment2, annotations = round(p.value, 3),
  y_position = y_position), # Position the annotations
  manual = TRUE,
  inherit.aes = FALSE
)
```
